---
title: 'Genetics graphing in R: integrating ggplot + StrataG'
author: "Lisa Komoroske"
date: '`r Sys.Date()`'
output: html_document
---

*N.B. Work in progress integrating with StrataG Pop Structure script and Eric's new code for making everything compatible with gtypes objects*

#Graphical Exploration of for genetic data

This document covers how to make some plots and manipulate data that are common to genetic analyses. This is a sister script to the population structure script using StrataG, so I've made the code compatible with gtypes objects and other StrataG syntax.


## Intro- loading libraries
First we need to load the libraries if they are not already
*You may need to install these if they are not already installed*
*insert link with instructions about how to install libraries in RStudio*
```{r haplo_load, echo=T}

rm(list = ls())
library(strataG)
library(ggplot2)
library(reshape2)
library(dplyr)
library(tidyr)
devtools::install_github('ericarcher/strataG', build_vignettes = TRUE)
```
*N.B. You will need to periodically update StrataG from Eric's github which has latest changes-this is what the last line of code above does (this requires an internet connection and devtools to already be installed for this to run correctly)*

##1. Haplotype Allelic Frequency Bar Charts

Then, let's generate data to play with:
```{r haplo_freq, echo=T}

# example data
set.seed(8) #set the seed number for the random number generator so results are reproducible
df <- data.frame(
  id = 1:40,#create sequence of numbers which identifies the samples
  location = rep(c('Mainland MU', 'Yakushima MU', 'Ryukyu MU', 'Muroto bycatch'), each = 10),#repeat command, create a list of site names, each 10 times
  hap = sample(c('CcP1.1','CcP2.1','CcP2.2','CcP2.3'#creates list of haplotypes
                 ), 40, rep = TRUE)#randomly pick from above list, 40 times
)
```

Then we can check the structure and summary of our data to make sure it is what we want:
```{r haplo_str, echo=T}
str(df) #examine structure of dataframe
summary(df) #summary stats of dataframe
head(df)#first five rows of dataframe
```

Once we've checked that our data is ok, we transform into a gtypes object.
This makes it compatible with analyses in StrataG.  
Alternatively, if you already have been doing StrataG analyses, you would skip generating the data, and just use your already existing gtypes object.
```{r haplo_gtpyes, echo=T}
df.g <- df2gtypes(df, ploidy = 1) # make gtypes object for haplotypic data (the 'ploidy' part-so if you have diploid data, would be ploidy=2)
```

This data is by individual, typically what we'd have in a raw data file. But we may want to instead have summarized data for different graphs or analyses. There are several ways to do this. First (and easiest), we can extract the frequencies and proportions from the gtypes object (via the existing StrataG syntax):

```{r haplo_StrataG_freq, echo=T}
# Haplotype frequencies:
hap.freqs <- alleleFreqs(df.g, by.strata = T)$hap
hap.freqs[, "freq", ]
hap.freqs[, "prop", ]
```

Alternatively (e.g., if you are working with other types of data/not in StrataG), we can use the 'count' function in dplyr to create a frequency table of our data:
```{r haplo_count, echo=T}
summary.data <- count(df, location, hap, sort=T) 
head(summary.data)
```

And then use layered functions that add a proportional variable:
```{r haplo_prop, echo=T}
summary.data2<-df %>%
  group_by(location,hap) %>% #define groups
  summarize(n=n()) %>%             #summarize counts the occurrences
  mutate(freq=n/sum(n))            #mutate creates a new variable that we call 'freq' that is the proportion given each group
head(summary.data2)#did it work how you intended?
```
******start here
Now we can use these summarized data to make stacked bar graphs


```{r haplo_stackedbar, echo=T}
# Plot haplotype frequencies
#1. Using StrataG/gtypes method:
freq.df <- melt(hap.freqs[, "freq", ])
colnames(freq.df) <- c("Haplotype", "Location", "Frequency")
ggplot(freq.df, aes(Location, Frequency, fill = Haplotype)) +
  geom_bar(stat = "identity")

prop.df <- melt(hap.freqs[, "prop", ])
colnames(prop.df) <- c("Haplotype", "Location", "Proportion")
ggplot(prop.df, aes(Location, Proportion, fill = Haplotype)) +
  geom_bar(stat = "identity")

#Can add/change aesthetics:
a <- ggplot(prop.df,
            aes(Location, Proportion, fill=Haplotype)) + 
  geom_bar(stat='identity')

b <- a + theme_bw() + 
  scale_fill_manual(values=c("blue","forestgreen","grey70","purple"))+
  theme(axis.title.x = element_text(face="bold", 
                                    colour="black", 
                                    size=20),
        axis.text.x  = element_text(angle=30, 
                                    vjust=0.5, 
                                    size=14),
        axis.title.y = element_text(face="bold", 
                                    colour="black", 
                                    size=20),
        axis.text.y  = element_text(angle=90, 
                                    vjust=0.5, 
                                    size=14)) +
  guides(fill=guide_legend(title="Haplotype")) + 
  xlab("Location") + ylab("Proportion")
print(b)

#2. If used alternate (dplyr code), non-genetic data method, same thing:
ggplot(summary.data, 
       aes(x=location,y=n, fill=hap)) + 
  geom_bar(stat='identity') #counts

ggplot(summary.data2, 
       aes(x=location,y=freq, fill=hap)) + 
  geom_bar(stat='identity')#proportions

```
The difference between plotting the counts vs. the proportions doesn't seem that important here, because we have roughly the same sample sizes among groups. But, this can really change the way the data looks if that is not the case so it may be more desirable to show one or the other form.

Also, if you had haplotype frequencies, and need to convert the other way to individuals for StrataG, etc., you could convert:
```{r haplo_freq_ind, echo=T}
freq.df <- freq2GenData(hap.freqs[, "freq", ])#this currently doesnt run, but I think I need to update 
freq.g <- df2gtypes(freq.df, ploidy = 1)
```


Now, let's try using some realish data for an extended exercise
```{r genetic_real, echo=T}

mtdna <- read.csv('data/mtDNA_freq.csv')
str(mtdna) #data is in wide format

mtdna.long<-gather(mtdna,"haplotype","n", 2:9) #'gather' converts wide to long format

a<-ggplot(mtdna.long, 
       aes(x=Location,y=n, fill=haplotype)) + 
  geom_bar(stat='identity') +
  scale_fill_brewer(palette="Spectral") + 
  theme_bw()#play around with the themes and color scales as you want...
b <- a + theme_bw() + 
  theme(axis.title.x = element_text(face="bold", 
                                    colour="black", 
                                    size=16),
        axis.text.x  = element_text(angle=30, 
                                    vjust=0.5, 
                                    size=12),
        axis.title.y = element_text(face="bold", 
                                    colour="black", 
                                    size=16),
        axis.text.y  = element_text(angle=0, 
                                    vjust=0.5, 
                                    size=12)) +
  guides(fill=guide_legend(title="Haplotype")) + 
  xlab("Location") + ylab("Count")
print(b)
```



Next, we may want to make bar charts with error bars for Mixed Stock Analyses (MSA) results.  

In ggplot, you can change the formatting to be almost anything you want, but this can sometimes make the code long and more complex. So if there's an error, it can be challenging to find the source when it's all lumped together. To help with this, I tend to follow the structure of first setting up my plot with the basics to make sure it's working correctly with my data, and then I layer the formatting I want on top of that. 

So let's first do a simple example.
```{r mixed_stock, echo = T}
MSA<-read.csv("data/MSA_Ccar.csv")
str(MSA)
bar1<-ggplot(MSA, 
             aes(x=Nesting.Stock, y=Mean, fill=Weight ))  +
  geom_bar(colour="black",
           stat="identity", 
           position=position_dodge(), 
           width=0.7)+ 
  theme_bw()+  
  geom_errorbar(aes(ymin=Mean-SD, ymax=Mean+SD), 
                width=0, position=position_dodge(0.7))

bar2<-bar1+ theme(axis.title.x = element_text(face="bold", 
                                              colour="black", 
                                              size=16),
                  axis.text.x  = element_text(angle=0, 
                                              vjust=0.5, 
                                              size=12),
                  axis.title.y = element_text(face="bold", 
                                              colour="black", 
                                              size=16),
                  axis.text.y  = element_text(angle=90, 
                                              vjust=0.5, 
                                              size=12),
                  legend.title = element_blank(), 
                  legend.position=c(0.07,0.95)) + 
  xlab("Nesting Stock")+ylab("Estimated Contribution")+
  scale_fill_manual(values=c("blue","forestgreen"))
print(bar2) #change colors, legend positioning etc as desired...
```

Now make your own, slight more complex MSA plot (MJ's data)
```{r MJanalysis, echo=TRUE}

MJ<-read.csv("data/MJ_models_combined.csv", header=TRUE)
str(MJ)

bar1<-ggplot(MJ, aes(x=STOCK,y=MEAN, fill=Model))  +
  geom_bar(colour="black", 
           stat="identity",
           position=position_dodge(),
           width=0.7)+ 
  theme_bw()+  
  geom_errorbar(aes(ymin=MEAN-SD, ymax=MEAN+SD), width=0, 
                position=position_dodge(0.7))

bar2<-bar1 + theme(axis.title.x = element_text(face="bold", 
                                               colour="black", 
                                               size=18),
                  axis.text.x  = element_text(angle=70, 
                                              vjust=0.5, 
                                              size=14),
                  axis.title.y = element_text(face="bold", 
                                              colour="black", 
                                              size=18),
                  axis.text.y  = element_text(angle=90, 
                                              vjust=0.5, 
                                              size=14),
                  legend.title=element_blank(), 
                  legend.position=c(0.9,0.85))+
  xlab("Nesting Stock")+ylab("Estimated Contribution") + 
  scale_fill_manual(values=c("blue","forestgreen"))+
  coord_cartesian(ylim=c(-0.01,0.78))#sets the axis limits
print(bar2)

#Can also save the figure directly to file:
#ppi=300 #define a pixels per inch term
#png("Figure1.png", width=10*ppi, height=6*ppi, res=ppi) #define your png file
#bar2 #create the plot
#dev.off() #stop!
```

Finally, we may want to look at the Allele Frequencies as part of our data exploration. When we have multiple loci with many alleles (e.g., microsatellites), it can be laborious to make the graphs one by one. Faceting is a great option for this scenario, because it allows you to quickly make groups of graphs by a certain factor. Here I use the same style of input data as you would have prepared for StrataG analyses.
```{r MSat_AlleleFreq, echo=T}
#3. MSat Allele Freq graph####
Msat.Ei.geno <- read.csv("data/EiMsatData020915.csv") 
Msat.Ei.strata <- read.csv("data/Ei_Msat_strata.csv") #LABID and population from analysis set.
Msat_Ei<-merge(Msat.Ei.geno,Msat.Ei.strata)
Msat_Ei_subset<-Msat_Ei[,c(52,2:13)]
head(Msat_Ei_subset)
Ei_long<-gather(Msat_Ei_subset,"Locus","allele_ID",2:13)
head(Ei_long)
Ei_long2<-Ei_long%>%
  separate(Locus,into=c("Locus","allele"),sep="_")
head(Ei_long2)
hist<-ggplot(Ei_long2,aes(x=allele_ID,fill=Population))+
  geom_histogram(binwidth = 5)+theme_bw()+
  theme(axis.text.x  = element_text(angle=70, vjust=0.5,size=10))
hist#this puts all our Locus together-not what we want
hist2<-hist+facet_wrap(~Locus,scales="free")#the scales=free is important when have different ranges among groups (as is the case with Msats, etc.)
#quartz(10,10)#this is for Mac, just pops out a big external graphing window-I find this useful when using two monitors; there is an analagous command for PC
print(hist2)

#if want to write to file-
#ppi=300 #define a pixels per inch term
#png("Figure2.png", 
#    width=20*ppi, 
#    height=12*ppi, 
#    res=ppi) #define your png file
#hist2
#dev.off()
```

 ###Mean Heterozygosity by locus
```{r graphics, include=TRUE}
data<-read.csv("EiMSat_2017-02-03_Ho.csv")#change accordingly
data<-rename(data, c("X"="locus", "x"="mean.heterozygosity"))
head(data)
#plot the mean.heterozygosity####
pdf(paste(description, "_mean_Het_by_locus", ".pdf", sep = ""), width = 15, height = 8)

ggplot(data = data, aes(x = locus, y = mean.heterozygosity)) + geom_point(size=4, shape = 18, fill = "black") + 
  labs(x = "Locus", y = "Observed Heterozygosity\n") + theme_classic()
dev.off()
```

#Bonus-add in later: mapping
Show graph of Haplotype Freqs overlaid on map from tutorial from K. Gilbert Could be for future workshop as people advance if people are interested


