---
title: 'Genetics graphing in R: integrating ggplot + StrataG'
author: "Lisa Komoroske"
date: '`r Sys.Date()`'
output: html_document
---

*N.B. Work in progress integrating with StrataG Pop Structure script and Eric's new code for making everything compatible with gtypes objects*

##Graphical Exploration of for genetic data  

This document covers how to make some plots and manipulate data that are common to genetic analyses. This is a sister script to the population structure script using StrataG, so for most sections I've made 2 options:  
(1) where code compatible with gtypes objects and other StrataG syntax
(2) where code is amenable to just normal R dataframe format

### Intro- loading libraries
First we need to load the libraries if they are not already
*You may need to install these if they are not already installed*
[Instructions to do this can be found here][1]
```{r haplo_load, echo=T}

rm(list = ls())
library(strataG)
library(ggplot2)
library(reshape2)
library(dplyr)
library(tidyr)
#devtools::install_github('ericarcher/strataG', build_vignettes = TRUE) uncomment and run if need to update StrataG
```
*N.B. You will need to periodically update StrataG from Eric's github which has latest changes-this is what the last line of code above does (this requires an internet connection and devtools to already be installed for this to run correctly)*

###Basic Graphics:  
####1. Haplotype Allelic Frequency Bar Charts

First, let's generate data to play with:
```{r haplo_freq, echo=T}
# example data
set.seed(8) #set the seed number for the random number generator so results are reproducible
df <- data.frame(
  id = 1:40,#create sequence of numbers which identifies the samples
  location = rep(c('Mainland MU', 'Yakushima MU', 'Ryukyu MU', 'Muroto bycatch'), each = 10),#repeat command, create a list of site names, each 10 times
  hap = sample(c('CcP1.1','CcP2.1','CcP2.2','CcP2.3'#creates list of haplotypes
                 ), 40, rep = TRUE)#randomly pick from above list, 40 times
)
```
(Obviously, here you can also load your real data here to use)  

Then we can check the structure and summary of our data to make sure it is what we want:  
```{r haplo_str, echo=T}
str(df) #examine structure of dataframe
summary(df) #summary stats of dataframe
head(df)#first five rows of dataframe
```

Once we've checked that our data is ok, we transform into a gtypes object. This makes it compatible with analyses in StrataG.  
```{r haplo_gtpyes, echo=T}
df.g <- df2gtypes(df, ploidy = 1) # make gtypes object for haplotypic data (the 'ploidy' part-so if you have diploid data, would be ploidy=2)
```
Alternatively, if you already have been doing StrataG analyses, you would skip steps above, and just use your already existing gtypes object.

So we see that these data are by individual, typically what we'd have in a raw data file. But we may want to instead have summarized data for different graphs or analyses. There are several ways to do this. 

####First (and easiest), we can extract the frequencies and proportions from the gtypes object (via the existing StrataG syntax):  
```{r haplo_StrataG_freq, echo=T}
# Extract Haplotype frequencies from gtypes object:
hap.freqs <- alleleFreqs(df.g, by.strata = T)$hap
hap.freqs[, "freq", ]
hap.freqs[, "prop", ]

#If you want to do the Reverse- Also, if you had haplotype frequencies, and need to convert the other way to individuals for StrataG, etc., you could convert by:
freq.df <- freq2GenData(hap.freqs[, "freq", ])
freq.g <- df2gtypes(freq.df, ploidy = 1)

# Now we can Plot haplotype frequencies: use these summarized data to make stacked bar graphs
freq.df <- melt(hap.freqs[, "freq", ])
colnames(freq.df) <- c("Haplotype", "Location", "Frequency")
ggplot(freq.df, aes(Location, Frequency, fill = Haplotype)) +
  geom_bar(stat = "identity")

prop.df <- melt(hap.freqs[, "prop", ])
colnames(prop.df) <- c("Haplotype", "Location", "Proportion")
ggplot(prop.df, aes(Location, Proportion, fill = Haplotype)) +
  geom_bar(stat = "identity")

#Can add/change aesthetics:
a <- ggplot(prop.df,
            aes(Location, Proportion, fill=Haplotype)) + 
  geom_bar(stat='identity')

b <- a + theme_bw() + 
  scale_fill_manual(values=c("blue","forestgreen","grey70","purple"))+
  theme(axis.title.x = element_text(face="bold", 
                                    colour="black", 
                                    size=20),
        axis.text.x  = element_text(angle=30, 
                                    vjust=0.5, 
                                    size=14),
        axis.title.y = element_text(face="bold", 
                                    colour="black", 
                                    size=20),
        axis.text.y  = element_text(angle=90, 
                                    vjust=0.5, 
                                    size=14)) +
  guides(fill=guide_legend(title="Haplotype")) + 
  xlab("Location") + ylab("Proportion")
print(b)
```
**Note:**-In ggplot, you can change the formatting to be almost anything you want, but this can sometimes make the code long and more complex. So if there's an error, it can be challenging to find the source when it's all lumped together. To help with this, I tend to follow the structure of first setting up my plot with the basics to make sure it's working correctly with my data, and then I layer the formatting I want on top of that.

####Secondly/Alternatively (e.g., if you are working with other types of data/not in StrataG), we can use the 'count' function in dplyr to create a frequency table of our data:  
```{r haplo_count_prop, echo=T}
summary.data <- count(df, location, hap, sort=T) 
head(summary.data)#did it work how you intended?
#And/or use layered functions and also add a proportional variable:
summary.data2<-df %>%
  group_by(location,hap) %>% #define groups
  summarize(n=n()) %>%             #summarize counts the occurrences
  mutate(freq=n/sum(n))            #mutate creates a new variable that we call 'freq' that is the proportion given each group
head(summary.data2)#did it work how you intended?

# Now we can Plot haplotype frequencies: use these summarized data to make stacked bar graphs (so this is the general code that works with any type of count/prop data, doesn't have to be genetic haplotypes):
ggplot(summary.data, 
       aes(x=location,y=n, fill=hap)) + 
  geom_bar(stat='identity') #counts

ggplot(summary.data2, 
       aes(x=location,y=freq, fill=hap)) + 
  geom_bar(stat='identity')#proportions

#use same code above for changing aesthetics to your liking
```
For more details on changing aesthetics in ggplot, see [Resources Here][2]   

**Note:** The difference between plotting the counts vs. the proportions doesn't seem that important here, because we have roughly the same sample sizes among groups. But, this can really change the way the data looks if that is not the case so it may be more desirable to show one or the other form.  

####Now, let's try using some realish data for an extended exercise
```{r geneticdata_real, echo=T}
mtdna <- read.csv('PGdata/mtDNA_freq.csv')
str(mtdna) #See that the data is in wide format

mtdna.long<-gather(mtdna,"haplotype","n", 2:9) #'gather' converts wide to long format (this works for the non-gtypes object graph example)

#If we need to create a StrataG gtypes object, we want each row to actually be one sample. Often our raw data is in that format (why we generated the data in the example that way) that then needs to get converted into a frequency table. But, perhaps you made and/or a coauthor sent you a summarized table like we have here and you need to reverse it to get into a gtypes object 

#(I know, this all seems circular, but what I am trying to provide are ways to easily move between data formats, no matter which format you start with)

#So, then we can use some frequency to binomial conversion code:
dataset<-mtdna.long
dataset.expanded = dataset[0,]
for (i in 1:length(dataset$n))
{
 if(dataset$n[i]>0)
 {
   dataset.add.succ = dataset[rep(i,dataset$n[i]),]
   dataset.add.succ$n=1
   #dataset.add.succ$failure=0
   dataset.expanded=rbind(dataset.expanded, dataset.add.succ)
 }
}
dataset.expanded$id <- 1:nrow(dataset.expanded)#just adding a dummy sample ID column since strataG needs that
mtdna.expanded<-dataset.expanded [,c(4,1:2)] #now each row is an observation
df.realg <- df2gtypes(mtdna.expanded, ploidy = 1) #make gtypes object
```

####Make stacked bar charts:  
#####1. StrataG/gtypes option:
```{r option1_realdata, echo=T}

# Extract Haplotype frequencies from gtypes object:
hap.freqs.real <- alleleFreqs(df.realg, by.strata = T)$hap
hap.freqs.real[, "freq", ] #this should look like our original input data (rows/columns reversed)-check, does it?
hap.freqs.real[, "prop", ]

freq.real.df <- freq2GenData(hap.freqs.real[, "freq", ])#reverse -to get individuals out of gtypes object if you started there (again, this shoudl be the same as the mtdna.expanded we created above!)

# Then Plot haplotype frequencies: use these summarized data to make stacked bar graphs
freq.real.df <- melt(hap.freqs.real[, "freq", ])
colnames(freq.real.df) <- c("Haplotype", "Location", "Frequency")
ggplot(freq.real.df, aes(Location, Frequency, fill = Haplotype)) +
  geom_bar(stat = "identity")

prop.real.df <- melt(hap.freqs.real[, "prop", ])
colnames(prop.real.df ) <- c("Haplotype", "Location", "Proportion")
ggplot(prop.real.df , aes(Location, Proportion, fill = Haplotype)) +
  geom_bar(stat = "identity")

#Can add/change aesthetics:
a <- ggplot(prop.real.df ,
            aes(Location, Proportion, fill=Haplotype)) + 
  geom_bar(stat='identity')

b <- a + theme_bw() + 
  scale_fill_manual(values=c("blue","forestgreen","grey70","purple","black","orange","yellow","cyan"))+#change to whatever colors float your boat
  theme(axis.title.x = element_text(face="bold", 
                                    colour="black", 
                                    size=20),
        axis.text.x  = element_text(angle=30, 
                                    vjust=0.5, 
                                    size=14),
        axis.title.y = element_text(face="bold", 
                                    colour="black", 
                                    size=20),
        axis.text.y  = element_text(angle=90, 
                                    vjust=0.5, 
                                    size=14)) +
  guides(fill=guide_legend(title="Haplotype")) + 
  xlab("Location") + ylab("Proportion")
print(b)
```

#####1. Non-StrataG/gtypes aka normal dataframe option:
```{r option2_realdata, echo=T}
a<-ggplot(mtdna.long, 
       aes(x=Location,y=n, fill=haplotype)) + 
  geom_bar(stat='identity') +
  scale_fill_brewer(palette="Spectral") + 
  theme_bw()#play around with the themes and color scales as you want...

b <- a + theme_bw() + 
  theme(axis.title.x = element_text(face="bold", 
                                    colour="black", 
                                    size=16),
        axis.text.x  = element_text(angle=30, 
                                    vjust=0.5, 
                                    size=12),
        axis.title.y = element_text(face="bold", 
                                    colour="black", 
                                    size=16),
        axis.text.y  = element_text(angle=0, 
                                    vjust=0.5, 
                                    size=12)) +
  guides(fill=guide_legend(title="Haplotype")) + 
  xlab("Location") + ylab("Count")
print(b)
```

####Next, we may want to look at the Allele Frequencies as part of our data exploration. 

When we have multiple loci with many alleles (e.g., microsatellites), it can be laborious to make the graphs one by one. Faceting is a great option for this scenario, because it allows you to quickly make groups of graphs by a certain factor. Here I again show two examples:
1. Using raw data in the same style of input data as you would have prepared for StrataG analyses  
2. Data already in a gtypes object-using data one of the vignettes from the StrataG package  
```{r MSat_AlleleFreq, echo=T}
#Option 1: Data already in gtypes object:
data(msats.g)#dolphin msat data from StrataG vignette 
msat.freqs <- alleleFreqs(msats.g, by.strata = TRUE)
msat.freqs <- melt(msat.freqs)
colnames(msat.freqs) <- c("allele", "type", "Strata", "value", "Locus")
msat.freqs$allele <- factor(msat.freqs$allele)

ggplot(subset(msat.freqs, type = "freq"), aes(allele, value, fill = Strata)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Locus, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    legend.title = element_blank()
  )

#Option 2: Raw data organized to go into StrataG:
Msat.Ei.geno <- read.csv("PGdata/EiMsatData020915.csv") 
Msat.Ei.strata <- read.csv("PGdata/Ei_Msat_strata.csv") #LABID and population from analysis set.
#Note: here could also circle back to StrataG analysis script and create gtypes object and then use code above
Msat_Ei<-merge(Msat.Ei.geno,Msat.Ei.strata)
Msat_Ei_subset<-Msat_Ei[,c(52,2:13)]
head(Msat_Ei_subset)
Ei_long<-gather(Msat_Ei_subset,"Locus","allele_ID",2:13)
head(Ei_long)
Ei_long2<-Ei_long%>%
  separate(Locus,into=c("Locus","allele"),sep="_")
head(Ei_long2)
hist<-ggplot(Ei_long2,aes(x=allele_ID,fill=Population))+
  geom_histogram(binwidth = 5)+theme_bw()+
  theme(axis.text.x  = element_text(angle=70, vjust=0.5,size=10))
hist#this puts all our Locus together-not what we want
hist2<-hist+facet_wrap(~Locus,scales="free")#the scales=free is important when have different ranges among groups (as is the case with Msats, etc.)
#quartz(10,10)#this is for Mac, just pops out a big external graphing window-I find this useful when using two monitors; there is an analagous command for PC
print(hist2)

#if want to write to file-
#ppi=300 #define a pixels per inch term
#png("Figure2.png", 
#    width=20*ppi, 
#    height=12*ppi, 
#    res=ppi) #define your png file
#hist2
#dev.off()
```


###Next, we may want to make bar charts with error bars for Mixed Stock Analyses (MSA) results.  

Here we are using summary data that would be generated from programs external to R (so not integrated with StrataG); although there may be increasing use of R packages for MSA, so at a later date I may come back and integrate data structure. For now, we'll read in summary data to generate the plots.

So again let's first start with a simple example.
```{r mixed_stock, echo = T}
MSA<-read.csv("PGdata/MSA_Ccar.csv")
str(MSA)
bar1<-ggplot(MSA, 
             aes(x=Nesting.Stock, y=Mean, fill=Weight ))  +
  geom_bar(colour="black",
           stat="identity", 
           position=position_dodge(), 
           width=0.7)+ 
  theme_bw()+  
  geom_errorbar(aes(ymin=Mean-SD, ymax=Mean+SD), 
                width=0, position=position_dodge(0.7))

bar2<-bar1+ theme(axis.title.x = element_text(face="bold", 
                                              colour="black", 
                                              size=16),
                  axis.text.x  = element_text(angle=0, 
                                              vjust=0.5, 
                                              size=12),
                  axis.title.y = element_text(face="bold", 
                                              colour="black", 
                                              size=16),
                  axis.text.y  = element_text(angle=90, 
                                              vjust=0.5, 
                                              size=12),
                  legend.title = element_blank(), 
                  legend.position=c(0.07,0.95)) + #change this according to where you want it
  xlab("Nesting Stock")+ylab("Estimated Contribution")+
  scale_fill_manual(values=c("blue","forestgreen"))
print(bar2) #change colors, legend positioning etc as desired...
```

Now make your own, slight more complex MSA plot  
```{r MJanalysis, echo=TRUE}
MJ<-read.csv("PGdata/MJ_models_combined.csv", header=TRUE)
str(MJ)

bar1<-ggplot(MJ, aes(x=STOCK,y=MEAN, fill=Model))  +
  geom_bar(colour="black", 
           stat="identity",
           position=position_dodge(),
           width=0.7)+ 
  theme_bw()+  
  geom_errorbar(aes(ymin=MEAN-SD, ymax=MEAN+SD), width=0, 
                position=position_dodge(0.7))

bar2<-bar1 + theme(axis.title.x = element_text(face="bold", 
                                               colour="black", 
                                               size=18),
                  axis.text.x  = element_text(angle=70, 
                                              vjust=0.5, 
                                              size=14),
                  axis.title.y = element_text(face="bold", 
                                              colour="black", 
                                              size=18),
                  axis.text.y  = element_text(angle=90, 
                                              vjust=0.5, 
                                              size=14),
                  legend.title=element_blank(), 
                  legend.position=c(0.9,0.85))+
  xlab("Nesting Stock")+ylab("Estimated Contribution") + 
  scale_fill_manual(values=c("blue","forestgreen"))+
  coord_cartesian(ylim=c(-0.01,0.78))#sets the axis limits
print(bar2)

#Can also save the figure directly to file:
#ppi=300 #define a pixels per inch term
#png("Figure1.png", width=10*ppi, height=6*ppi, res=ppi) #define your png file
#bar2 #create the plot
#dev.off() #stop!
```

Finally, of course these are just some examples of basic plots one might be interested in for genetic analyses-there are many others that might be of interest depending on system/question
e.g.,
 ###Mean Heterozygosity by locus
```{r graphics, include=TRUE}
data<-read.csv("PGdata/EiMSat_April_20_Ho.csv")#Heterozygosity data generated in StrataG analysis script-change file name accordingly
library(plyr)
data<-rename(data, c("X"="locus", "x"="mean.heterozygosity"))
head(data)
#plot the mean.heterozygosity####
#pdf(paste(description, "_mean_Het_by_locus", ".pdf", sep = ""), width = 15, height = 8)
ggplot(data = data, aes(x = locus, y = mean.heterozygosity)) + geom_point(size=6, shape = 23, fill = "red",alpha=0.5) + 
  labs(x = "Locus", y = "Observed Heterozygosity\n") + theme_classic()+  theme(axis.text.x  = element_text(angle=70, vjust=0.5,size=10))
#dev.off()
```

#Bonus-add in later: mapping
Show graph of Haplotype Freqs overlaid on map from [tutorial from K. Gilbert] [3] (Could be for future workshop as people advance if people are interested)

[1]:https://www.youtube.com/watch?v=u1r5XTqrCTQ
[2]: http://www.cookbook-r.com/Graphs/
[3]:http://www.molecularecologist.com/2012/09/making-maps-with-r/